Rapid Epidemiological Analysis of Comorbidities and Treatments in COVID-19 in Scotland (REACT-SCOT): a case-control study based on the Scottish population

### 2020-05-12 casecontrol.Rmd
Fixed to use refactored version

---
title: 'Risk factors for severe COVID-19 in Scotland: a population-based case-control study'
author: 
  - name: Paul M McKeigue
    email: paul.mckeigue@ed.ac.uk
    affiliation: Usher
  - name: Amanda Weir 
    affiliation: HPS
  - name: Jen Bishop
    affiliation: HPS
  - name: Stuart McGurnaghan
    affiliation: IGMM
  - name: Sharon Kennedy
    affiliation: HPS
  - name: Sharon Hutchinson 
    affiliation: GCU
  - name: David McAllister 
    affiliation: Glasgow
  - name: Chris Robertson 
    affiliation: Strathclyde
  - name: Rachael Wood
    affiliation: ISD
  - name: Nazir Lone
    affiliation: Usher
  - name: Tom Caparrotta 
    affiliation: IGMM
  - name: Helen M Colhoun
    affiliation: IGMM
address: 
   - code: Usher
     address: Usher Institute, College of Medicine and Veterinary Medicine, University of Edinburgh, Teviot Place, Edinburgh EH8 9AG, Scotland. PM - Professor of Genetic Epidemiology and Statistical Genetics. NL - Clinical Senior Lecturer in Critical Care
   - code: IGMM
     address: Institute of Genetics and Molecular Medicine, College of Medicine and Veterinary Medicine, University of Edinburgh, Western General Hospital Campus, Crewe Road, Edinburgh EH4 2XUC, Scotland. HC - Axa Chair in Medical Informatics and Epidemiology. TC - Sir George Alberti Doctoral Fellow in Pharmacoepidemiology. 
   - code: Glasgow
     address: Institute of Health and Wellbeing, University of Glasgow, 1 Lilybank Gardens, Glasgow G12 8RZ. DM - Senior Clinical Lecturer in Public Health 
   - code: HPS
     address: Health Protection Scotland (Public Health Scotland), Meridian Court, 5 Cadogan Street, Glasgow G2 6QE
   - code: Strathclyde
     address: Department of Mathematics and Statistics, University of Strathclyde, 16 Richmond Street, Glasgow G1 1XQ. CR - Professor of Public Health Epidemiology
   - code: GCU
     address: School of Health and Life Sciences, Glasgow Caledonian University. SH - Professor of Epidemiology and Population Health 
   - code: ISD
     address: NHS Information Services Division (Public Health Scotland), Gyle Square, 1 South Gyle Crescent, Edinburgh, EH12 9EB.  RW - Consultant in Maternal and Child Health.

geometry: margin=1in
fontsize: 11pt
linenumbers: true
bibliography: ../covid.bib
date: "This version compiled `r gsub('^0', '', format(Sys.time(), '%d %B %Y'))`"

header-includes: |
  \let\origquote\quote
  \def\quote{\origquote\itshape}
  \usepackage{graphicx}
  \usepackage{longtable}
  \usepackage{booktabs}
  \usepackage{float}
  \usepackage{array}
output: 
#  pdf_document:
  rticles::plos_article:
    latex_engine: lualatex
    fig_caption: true
    keep_tex: true
#    toc: true
#    toc_depth: 2
csl: ../plos.csl
urlcolor: red

@@

@@

@@

# Background
Reports based on cases seen in hospitals hve suggested that in those with severe disease the prevalence of diabetes and cardiovascular disease is higher than expected, and that a higher than expected proportion of severe cases are of non-European ethnic origin. Quantification of the risk associated with these characteristics and co-morbidities has been limited by the lack of an appropriate background population comparator. Furthermore it has not been possible to disentangle whether drug effects may underlie the apparent association with certain co-morbid groups.

The aim of this study was to investigate risk factors for severe or fatal cases of COVID-19 confirmed by testing.  Specific objectives were to test hypotheses about drugs classes associated with altered risk, and to construct a risk prediction algorithm that can be used to identify high-risk individuals who could be offered shielding during the epidemic.  To achieve this aim we conducted a case-control study of all severe and fatal cases in Scotland, with controls matched for age, sex and primary care practice.

# Methods
## Case definition 
The Electronic Communication of Surveillance in Scotland (ECOSS) database captures all virology testing in all NHS laboratories nationally, using the Community Health Index (CHI) identifier.  The CHI number is a unique identifier for everyone in the population, and is used in all care systems in Scotland.  All individuals testing positive for nucleic acid for SARS-CoV-2 up to `r format(max(cc.all$SPECDATE, na.rm=TRUE), "%d %B %Y")` were ascertained.  Admissions to critical care were obtained from the national audit database of the Scottish Intensive Care Society and Audit Group (SICSAG)ical care units (SICSAG). Hospital admissions from the time of testing were obtained from the RAPID-24 database which is updated each day. 
Death registrations up to `r format(max(cc.all$Date.Death, na.rm=TRUE), "%d %B %Y")`
and cause of death were obtained from linkage to the National Register of Scotland.

@@

By restricting the case definition to a group that combines those receiving critical care with those who died without receiving critical care, we ensure complete ascertainment of severe test-positive cases, whatever selection procedures may have been in place to exclude those assessed as unlikely to benefit from critical care.  This focus on severe or fatal disease is relevant to policy because decisions about control measures have been based on projections of the requirement for critical care.

In national mortality data, XX % of those certified with COVID-19 as underlying case of death had a positive test record.

## Matched controls
For each test-positive case, we ascertained ten matched controls of the same sex, one year age band and attending the same General Practice as each case alive on the same day as the date of test in the case. Note that this is an incidence density sampling design i.e. a control who becomes a case at a later point is retained in the control set. With this design, the odds ratios in a conditional logistic regression model are equivalent to rate ratios.

## Demographic data
Ethnicity of cases and controls was classified using a name classification algorithm [@lakha_name_2011] with the names in the CHI database.  This is able to distinguish South Asian and Chinese groups from the White group, but misclassifies Black-Caribbean people (a relatively small group in Scotland) as White.  For those with a hospital admission record, a separate coding of ethnicity was available using self-described categories based on those used in the national census.  Socioeconomic status was scored as the Scottish Index of Multiple Deprivation, based on the postcode in the CHI database.

## Morbidity and drug prescribing

All cases and Using the CHI identifier, records from ECOSS were linked to hospital discharge records in the Morbidity Record 01 (SMR01).  and

For all cases and controls, ICD-10 diagnostic codes were extracted from the last five years of hospital discharge records in the Scottish Morbidity Record 01 (SMR01, excluding records of discharges less than 25 days before testing positive for SARS-Cov-2. BNF drug codes were extracted from the last year of encashed prescriptions, excluding those encashed less than 15 days before testing positive for SARS-CoV-2.  Diagnosed cases of diabetes were identified through linkage to the national diabetes register (SCI-Diabetes), with a clinical classification of diabetes type as Type 1, Type 2 or Other.

We initially examined history of a specific list of conditions and drug classes that have been advised as high risk conditions for COVID-19 by Public Health England, World Health organisation or the US Centres for Disease Control.

Co-morbid conditions were assigned based on either a diagnostic code in any hospital discharge record during the last five years, or an encashed prescription of a drug for which that diagnostic code is the only indication.

The ICD-10 codes and drug codes used to define each variable are listed in Supplementary Materials.

## Statistical methods

We calculated the sex and age band specific cumulative incidence of severe COVID-19 disease using the age and sex distribution of those alive in the population as of March 1st 2020. (CHECK)

The univariate association of each variable with severe COVID-19 was modelled with conditional logistic regression, implemented as Cox regression in the R function `survival::clogit`

We tested each effect for interaction with age and sex.  Effect of interest

The analyses were repeated after excluding those aged over 75 years.

To construct a risk prediction model, we used stepwise regression alternating between forward and backward steps, implemented in the R function `stats::step`, so as to maximize the AIC.

The performance of the risk prediction model as a classifier was examined by n-fold cross-validation, with performance calculated over all test folds as the expected information for discrimination and (more conventionally) as the C-statistic.

### Comorbidities of initial interest

- Circulatory Disorders
  - Ischaemic heart disease
  - Other heart disease
  - Cerebrovascular Disease
  - Other circulatory system diseases
- Neurological diseases except inflammatory
  - Epilepsy
  - Mono and  polyneuropathies
  - Other neurological conditions
- Respiratory diseases
  -  Acute respiratory infections
  -  Asthma
  -  Other Chronic lower respiratory disease
- Tuberculosis 
- Connective tissue diseases
- Decompensated liver disease
- Kidney disease
  - Advanced Chronic kidney disease or RRT
  - Other chronic kidney disease
- Diabetes
  - type 1
  - type 2
- Malignant neoplasms
  - Lung cancers
  - Blood cancers
  - Other cancers
- Immunological diseases
  - HIV
  - Certain disorders involving the immune system
- Sickle cell disease
- Cystic fibrosis
- Organ transplantation other than kidney
- Treated hypertension (on any antihypertensive drugs in past year)
- Charlson comorbidity index

​
### Other recent health care exposure variables of interest are

- Any inpatient admission in past year and excluding the three weeks prior to testing 
- Any emergency admission in past year and excluding the three weeks prior to testing
- Any ICU/HDU/Admission in past year and excluding the three weeks prior to testing

### Initial drug exposures of interest are

- Any lipid  drug 
- Any anticoag or platelet  
- Any Antihypertensive drugs disaggregated to: 
  - Vasodilator antihypertensive drugs
  - Centrally-acting antihypertensive drugs
  - Adrenergic neurone blocking drugs
  - Alpha-adrenoceptor blocking drugs
  - Angiotensin-converting enzyme inhibitors
  - Angiotensin-II receptor antagonists
  - Renin inhibitors
  - Thiazides and related diuretics
  - Calcium-channel blocker
- Any prescribed NSAID
- Any cytotoxic or immunosuppressant drug 
- Any chloroquine drug or derivatives of chloroquine

Derivations ICD10 and BNF codes for these comorbidities / drugs are in the appendix

Further comorbidities will be included in future protocol versions

Note that at this point we are not including high risk of exposure as covariates ie HCW, residential care home, prisoner etc this will form part of future work.

@@

# Results
## Comparison of severe cases with others who tested positive 
We emphasize that because testing was used selectively over this period, those who tested positive and were never hospitalised are unlikely to be representative of never-hospitalized cases in the population.

## Association of demographic factors with severe disease

For the case-control comparison, this version of the manuscript restricts the dataset to those aged less than `r max(cc.severe$AGE) + 1` years.

Table \ref{tab:demog} shows univariate associations of demographic factors with disease.  Higher risk of severe disease was associated with socioeconomic deprivation and with residence in a care home.  In comparison with those classified as White, South Asians and other non-White ethnic minorities were at lower risk of severe disease.

```{r demog, echo=FALSE, warning=FALSE, message=FALSE}
pvalue <- pvalue.latex(univariate.demog[, 5])
ci <- or.ci(univariate.demog[, 1], univariate.demog[, 3]) 
table.demog.aug <- data.frame(table.demog,
                                   ci, 
                                   pvalue)
colnames(table.demog.aug) <- c("Controls", "Cases", "Rate ratio (95% CI)", "p-value")
rownames(table.demog.aug) <- replace.names(rownames(table.demog.aug))

pander(table.demog.aug, 
       digits=c(0, 0, 3, 3, 3, 1),
       round=c(0, 0, 3, 3, 3, 10),
	   split.tables="Inf",
       align=rep("centre", 4), 
       split.cells=c(16, 16, 25, 8),
       caption="\\label{tab:demog}Univariate associations of severe disease with demographic factors")

## Associations of comorbid conditions with severe disease
In univariate logistic regression models, both Type 1 and Type 2 diabetes were associated with severe disease (Table \ref{tab:conditions}. There were strong associations with circulatory disease, respiratory disease, kidney disease and neurological disease.  No association of severe disease with cystic fibrosis was detected.  A possible explanation for this may be that this group was effectively shielded from infection in accordance with official guidance.

pvalue <- pvalue.latex(univariate.conditions[, 5])
ci <- or.ci(univariate.conditions[, 1], univariate.conditions[, 3]) 
table.conditions.aug <- data.frame(table.conditions,
                                   ci, 
                                   pvalue)
colnames(table.conditions.aug) <- c("Controls", "Cases", "Rate ratio (95% CI)", "p-value")
rownames(table.conditions.aug) <- replace.names(rownames(table.conditions.aug))

pander(table.conditions.aug, 
       digits=c(0, 0, 3, 3, 3, 1),
       round=c(0, 0, 3, 3, 3, 10),
	   split.tables="Inf",
       split.cells=c(20, 20, 20, 7),
       caption="\\label{tab:conditions}Univariate association of severe disease with comorbid conditions")

Further analyses to adjust for co-morbidity ...

### Counts of co-morbid conditions

Table xx shows the distribution of numbers of the co-morbid conditions listed by WHO or CDC or PHE by age band in severe cases versus their controls. 
Across each age band there were people with severe COVID-19 who did not appear to have any of these co-morbid conditions.

## Associations of prescribed drugs with severe disease 
Table \ref{tab:drugs} shows that severe disease was associated with use of antihypertensive agents, anticoagulants, and lipid-lowering agents.  Use of ACE inhibitors or angiotensin receptor blockers was not associated with higher risk than use of other classes of antihypertensive agent.  Use of non-steroidal anti-inflammatory agents was weakly associated with severe disease.  Use of hydroxychloroquine was associated with severe disease.

Further analyses to adjust for co-morbidity ...

pvalue <- pvalue.latex(univariate.drugs[, 5])
ci <- or.ci(univariate.drugs[, 1], univariate.drugs[, 3]) 
table.drugs.aug <- data.frame(table.drugs,
                                   ci, 
                                   pvalue)
colnames(table.drugs.aug) <- c("Controls", "Cases", "Rate ratio (95% CI)", "p-value")
rownames(table.drugs.aug) <- replace.names(rownames(table.drugs.aug))

pander(table.drugs.aug, 
       digits=c(0, 0, 3, 3, 3, 1),
       round=c(0, 0, 3, 3, 3, 10),
	   split.tables="Inf",
       split.cells=c(20, 20, 20, 7), 
       caption="\\label{tab:drugs}Univariate association of severe disease with drugs prescribed in last year")

## Examining those with no apparent co-morbid conditions

A more in depth focus on severe cases at younger ages

As shown in table  among severe case below age 50  xxx% had either xxx or xxx or xxx but in xxx% none of these comorbid conditions were found. Among severe case below age 50-70  xxx% had either xxx or xxx or xxx but in xxx% none of these comorbid conditions were found.  However in those with none of these co morbid conditions xx% were on   drugs for xxx xxx and xxx.

## Multivariate analysis of risk factors for severe disease

A stepwise regression procedure was carried out with all demographic, morbidity and drug exposure variables, using alternating forward-backward selection.  The `r nrow(stepwise.all)` variables retained are shown in Table \ref{tab:stepwise}

table.stepwise <- as.data.frame(stepwise.all[, c(1, 5)])
table.stepwise[, 1] <- exp(table.stepwise[, 1])
colnames(table.stepwise) <- c("Rate ratio", "p.value")
rownames(table.stepwise) <- replace.names(rownames(table.stepwise))

table.stepwise$p.value <- pvalue.latex(table.stepwise$p.value)

pander(table.stepwise, table.style="multiline",
       # split.cells=c(5, 5, 5, 5, 5, 5, 5), 
	   digits=c(3, 1), 
	   split.table="Inf",
       caption="\\label{tab:stepwise}Stepwise regression: variables retained in model for severe disease")

## Predictive performance of a risk classifier
The predictive performance of the model chosen by stepwise regression was estimated by `r nfold`-fold cross-validation, and comparing observed and predicted case status over all test folds.  Table \ref{tab:wevid} shows that the C-statistic was `r summary(all.densities)[3]` and the expected information for discrimination was  `r summary(all.densities)[5]` bits.

show incremental comparison of different models ...

pander(summary(all.densities), table.style="multiline",
       split.cells=c(5, 5, 5, 5, 5, 5, 5), split.table="Inf",
       caption="\\label{tab:wevid}Prediction of severe COVID-19: cross-validation of model chosen by stepwise regression" )

```{r plotwdists, echo=FALSE, fig.cap="Cross-validation of model chosen by stepwise regression: distributions of weight of evidence", out.width="0.8\\textwidth"}

plotWdists(all.densities) + theme_grey(base_size = 18) + 
	theme(legend.title=element_blank())

# Discussion

What is the explanation of the 14-fold higher incidence of severe disease in those who have encashed at least one prescription in the last year?

>>>>

---
title: 'Rapid Epidemiological Analysis of Comorbidities and Treatments in COVID-19 in Scotland (REACT-SCOT): a case-control study based on the Scottish population'
author: 
  - name: Paul M McKeigue
    email: paul.mckeigue@ed.ac.uk
    affiliation: Usher
  - name: Amanda Weir 
    affiliation: HPS
  - name: Jen Bishop
    affiliation: HPS
  - name: Stuart McGurnaghan
    affiliation: IGMM
  - name: Sharon Kennedy
    affiliation: HPS
  - name: David McAllister 
    affiliation: Glasgow
  - name: Chris Robertson 
    affiliation: Strathclyde
  - name: Rachael Wood
    affiliation: ISD
  - name: Nazir Lone
    affiliation: Usher
  - name: Janet Murray
    affiliation: HPS 
  - name: Tom Caparrotta 
    affiliation: IGMM
  - name: Sharon Hutchinson 
    affiliation: GCU
  - name: Helen M Colhoun
    affiliation: IGMM
address: 
   - code: Usher
     address: Usher Institute, College of Medicine and Veterinary Medicine, University of Edinburgh, Teviot Place, Edinburgh EH8 9AG, Scotland. PM - Professor of Genetic Epidemiology and Statistical Genetics. NL - Clinical Senior Lecturer in Critical Care
   - code: IGMM
     address: Institute of Genetics and Molecular Medicine, College of Medicine and Veterinary Medicine, University of Edinburgh, Western General Hospital Campus, Crewe Road, Edinburgh EH4 2XUC, Scotland. HC - Axa Chair in Medical Informatics and Epidemiology. TC - Sir George Alberti Doctoral Fellow in Pharmacoepidemiology. 
   - code: Glasgow
     address: Institute of Health and Wellbeing, University of Glasgow, 1 Lilybank Gardens, Glasgow G12 8RZ. DM - Senior Clinical Lecturer in Public Health 
   - code: HPS
     address: Health Protection Scotland (Public Health Scotland), Meridian Court, 5 Cadogan Street, Glasgow G2 6QE
   - code: Strathclyde
     address: Department of Mathematics and Statistics, University of Strathclyde, 16 Richmond Street, Glasgow G1 1XQ. CR - Professor of Public Health Epidemiology
   - code: GCU
     address: School of Health and Life Sciences, Glasgow Caledonian University. SH - Professor of Epidemiology and Population Health 
   - code: ISD
     address: NHS Information Services Division (Public Health Scotland), Gyle Square, 1 South Gyle Crescent, Edinburgh, EH12 9EB.  RW - Consultant in Maternal and Child Health. 
geometry: margin=1in
fontsize: 11pt
linenumbers: true
bibliography: ../covid.bib
date: "This version compiled `r gsub('^0', '', format(Sys.time(), '%d %B %Y'))`"
header-includes: |
  \let\origquote\quote
  \def\quote{\origquote\itshape}
  \usepackage{graphicx}
  \usepackage{longtable}
  \usepackage{booktabs}
  \usepackage{float}
  \usepackage{array}
  \usepackage{booktabs}
  \usepackage{longtable}
  \usepackage{lscape}
output: 
#  pdf_document:
  rticles::plos_article:
    latex_engine: lualatex
    fig_caption: true
    keep_tex: true
#    toc: true
#    toc_depth: 2
csl: ../plos.csl
urlcolor: red

@@

knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)

library(knitr)
library(kableExtra)

table.densities <- rbind(summary(listed.densities),
                         summary(full.densities))[, 1:6]

@@

# Abstract

_Background_-- Why some people infected with Sars-CoV-2 develop severe COVID-19 disease and others do not is poorly understood.  Here we  used sociodemographic data and extensive previous hospitalisation and prescribing history among all people who have developed severe COVID-19 disease in Scotland compared to the background population to understand factors associated with severity.

_Methods_ -- We linked the records of those tested positive for SARS-C0V-2 viral RNA in the national database that captures virology testing in all NHS laboratories, to  hospital admissions, intensive care and mortality data. Severe COVID-19 was defined as disease requiring intensive care support or associated with death within 28 days of a positive test. We selected 10 sex, age and general practice based controls per test-positive case. For all cases and controls we extracted the International Classification of Disease (ICD)-10 codes from the previous five years of hospital discharge records  and the British National Formulary (BNF)  drug codes of  encashed  primary care issued prescriptions in the past year. We evaluated the risk of severe COVID-19 associated with  a pre-defined  set of  conditions using conditional logistic regression. We examined how much predictive performance  for severe COVID-19  a risk classifier built  and tested with four-fold cross validation based on these conditions could achieve and whether this could be improved using information from the rest of the hospitalisation and prescribing data.

_Findings_ -- In a logistic regression based on the age-sex distribution of the national population, the odds ratios were `r round(exp(logistic.coeffs[3, 1]), 2)` for a 10-year increase in age and `r round(exp(-logistic.coeffs[2, 1]), 2)` for male sex.  In the case-control analysis, the odds ratios were

`r table.agegr.all[1, 3]` for residence in a care home status, 
`r table.agegr.all[4, 3]` for diabetes, 
`r table.agegr.all[8, 3]` for chronic renal failure, 
`r table.agegr.all[5, 3]` for ischemic heart disease
`r table.agegr.all[9, 3]` for neurological disease, 
`r table.agegr.all[10, 3]` for chronic liver disease and 
`r table.agegr.all[11, 3]` for immune deficiency or suppresssion

Overall but INSERT xx% of severe cases versus INSERT xx% of controls had at least one of these INSERT ( xx versus xx% in those aged under 40 years).  The odds ratio for severe disease associated with encashment of at least one prescription drug in the past year was `r table.agegr.all[2, 3]` and the odds ratio associated with at least one hospital admission in the past five years was `r table.agegr.all[3, 3]`).  These associations with severe disease were present in those without any of the listed conditions, and were distributed over most BNF chapters and the associations with hospital diagnoses over most ICD-10 chapters.

In comparison with a baseline model based on care home residence only, adding listed conditions increased the information for discrimination to  INSERT 0.9 bits. Adding indicator variables for ICD-10 and BNF chapters increased the information for discrimination further to  INSERT 1.5 bits. From a model based on age and sex alone, the expected information for discrimination was `r round(cases.Lambda.agesex, 2)` bits.

*Conclusions* --Along with older age and male sex, severe COVID-19 disease is strongly associated with past medical history and the vast majority of younger persons developing severe disease have an underlying medical condition. The comorbidities associated with increased risk are not limited to the current lists of risk conditions highlighted by public health agencies.  A risk classifier that uses all the information available in medical and prescribing records, rather than only a limited set of conditions, will more accurately discriminate between low-risk and high-risk individuals who may require shielding until the SARS-CoV-2 epidemic is over.   The strong association with prescribing history we have observed may reflect co-morbidity or adverse drug effects and we are investigating this further.

@@

# Background
It is clear that  there is wide susceptibility to risk of severe COVID-19 disease  with older age being a predominant factor. Case series from many countries have  suggested that in those with severe disease the prevalence of diabetes and cardiovascular disease is higher than expected.  However there are also many anecdotal reports of apparently healthy young persons succumbing to disease.  Apart from co-morbidities the potential  role of ethnicity in disease severity has been highlighted.  However quantification of the risk associated with these characteristics and co-morbidities has been limited by the lack of an appropriate background population comparator.  In particular a  systematic evaluation of the past medical history in those with and without severe disease has not been reported.

Here we  used extensive health care record and prescribing history among all people who have developed severe COVID-19 disease in Scotland compared to the background population to understand sociodemographic and comorbidity factors associated with severe COVID -19 disease.  We defined  severe disease as viral RNA-confirmed diagnosis of COVID-19 as  that required critical care support or was associated with death within 28 days of testing.   We focused on severe disease it is of most  concern but also  because it can be completely ascertained  whereas  testing policies means that   complete ascertainment of less severe disease is not possible in most countries.

# Methods
## Case definition 
The Electronic Communication of Surveillance in Scotland (ECOSS) database captures all virology testing in all NHS laboratories nationally, using the Community Health Index (CHI) identifier.  The CHI number is a unique identifier for everyone in the population, and is used in all care systems in Scotland.  All individuals testing positive for nucleic acid for SARS-CoV-2 up to `r format(max(cc.all$SPECDATE, na.rm=TRUE), "%d %B %Y")` were ascertained.   Hospital admissions from the time of testing were obtained from the RAPID-24 database which is updated each day and the weekly updated Scottish Morbidity Record 01. Admissions to critical care were obtained from the Scottish Intensive Care Society and Audit Group (SICSAG)  database that captures admission to all critical care units across Scotland. Death registrations up to `r format(max(cc.all$Date.Death, na.rm=TRUE), "%d %B %Y")`
and cause of death were obtained from linkage to the National Register of Scotland.

@@

By restricting the case definition to a group that combines those receiving critical care with those who died without receiving critical care, we ensure complete ascertainment of severe test-positive cases, whatever selection procedures may have been in place to exclude those assessed as unlikely to benefit from critical care.

## Matched controls
For each test-positive case, we ascertained ten matched controls of the same sex, one year age band and attending the same General Practice as each case alive on the same day as the date of test in the case using the Community Health Index (CHI) database. Note that this is an incidence density sampling design i.e. a control who becomes a case at a later point is correctly retained in the control set. With this design, the odds ratios in a conditional logistic regression model are equivalent to rate ratios.

## Demographic data
Socioeconomic status was scored using the postcode based indicator  the Scottish Index of Multiple Deprivation using the postcode in the CHI database.  Ethnic group was captured in two ways: self designated ethnicity  was derived from past records in the hospital admissions and outpatient attendances  (Scottish Morbidity Records) using  categories based on those used in the national census. However since this is incomplete  we also used the a name classification algorithm ONOMAP [@lakha_name_2011] with the names in the CHI database.  This is able to distinguish South Asian and Chinese groups but not Black-Caribbean people from the White group.  Residential care home status was available via the CHI database.

## Morbidity and drug prescribing
For all cases and controls, ICD-10 diagnostic codes were extracted from the last five years of hospital discharge records in the Scottish Morbidity Record 01 (SMR01), excluding records of discharges less than 25 days before testing positive for SARS-Cov-2 and using all codes on the discharge. British National Formulary  (BNF) drug codes were extracted from the last year of encashed prescriptions, excluding those encashed less than 25 days before testing positive for SARS-CoV-2.  Diagnosed cases of diabetes were identified through linkage to the national diabetes register (SCI-Diabetes), with a clinical classification of diabetes type as Type 1, Type 2 or Other.

We first extracted history of a specific list of conditions and drug classes that have been  included as  risk conditions for COVID-19 by  public health agencies https://www.nhs.uk/conditions/coronavirus-covid-19/people-at-higher-risk-from-coronavirus/.  These co-morbid conditions were assigned based on either a diagnostic code in any hospital discharge record during the last five years, or an encashed prescription of a drug for which that diagnostic code is the only indication.  The ICD-10 codes and drug codes used to define each variable are listed in Supplementary Materials.  In addition to these pre-specified groupings of codes we then undertook a systematic evaluation of the association of all ICD-10 codes  and BNF  codes to chapter level.  We did not  specifically  tabulate  the conditions that have been listed as making people " clinically extremely vulnerable"  such as cystic fibrosis or leukaemia  since there is no  clinical equipoise  regarding risk in such persons and since they will be underrepresented in the case data as they have been in complete shielding since early in the epidemic.

## Statistical methods
We calculated the sex and age band specific cumulative incidence of severe COVID-19 disease using the age and sex distribution of those alive in the population as of March 1st 2020. (CHECK)  The univariate association of  sociodemographic variables, pre-defined conditions and discharge and prescribing history  with severe COVID-19 were modelled with conditional logistic regression, implemented as Cox regression in the R function `survival::clogit` We examined the associations by age band and overall.  Among those  cases and controls without any of the pre-defined conditions we then further examined associations of ICD-10 and BNF  chapter with severe COVID-19. To construct a risk prediction models, we used stepwise regression alternating between forward and backward steps, implemented in the R function `stats::step`, so as to maximize the AIC.  The performance of the risk prediction model as a classifier was examined by 4-fold cross-validation, with performance calculated over all test folds as the expected information for discrimination [@mckeigue_quantifying_2019] and (more conventionally) as the C-statistic.

@@

# Results
<!---
Still needed

- A statement on "In national mortality data, the n and % of those certified with COVID-19 as underlying case of death had/ did not have a positive test record"  or sensivitivty analysis if sharon provides a case control set of the deaths with ue07 code 
- Figure - ?? graph of population attributable risk fraction of listed conditions by age ?? not sure if this is worth the trouble ...
-  Run table 2 analyses for all subjects combined 
- what is currently tables 3,4,5  and 7  could be put into appendix and the table reference in text changed as appropriate ?
- tables currently 9,11,12 should be removed as will go in pharmacoepi  paper  --->
## Incidence and mortality from severe COVID-19 in the Scottish population
Logistic regression models were fitted to the proportions of cases and non-cases in the Scottish population, using the estimated population of Scotland in mid-year 2019 which are available by one-year age group up to age 90 years.  Figure \ref{fig:rates} shows the relationships of incidence and mortality rates to age for each sex separately.  The relationship of mortality to age was almost exactly linear on a logit scale.  For fatal and non-fatal cases combined, the mortality differentials between males and females widened in middle age before narrowing again after about age 75 years.  In models including both age and sex as covariates, the odds ratio associated with a 10-year increase in age was `r round(exp(logistic.coeffs[3, 1]), 2)` for all severe disease and `r round(exp(logistic.coeffs[3, 2]), 2)` for fatal disease.  The odds ratio associated  with male sex was is INSERT  for all severe disease and  INSERT  for fatal disease.  As shown in Table INSERT  almost all of the severe cases  death over aged 75years died  but  the majority of severe cases under aged 60 survived.

## Association with demographic factors
Table \ref{tab:demog} shows univariate associations of demographic factors with severe disease.  Higher risk of severe disease was associated with socioeconomic deprivation and with residence in a care home.  In comparison with those classified as White, South Asians and other non-White ethnic minorities were at lower risk of severe disease.  Table \ref{tab:ethnicsmr} shows a separate analysis restricted to those with ethnicity recorded in hospital records, which allows those who identify as Black to be distinguished from those who identify as White.  Although the numbers are small, there is no evidence that the Black minority is at higher risk of severe disease.

## Associations with listed conditions
Overall  INSERT xx of cases and xx of controls had at least one of the listed conditions.   The  prevalence of the listed conditions in cases and controls by age band is shown in Table  INSERT .   Almost half of cases aged under 40 years had at least one listed condition, compared with only 9% (43 / 470) of controls.  In those 75+ years INSERT xxx 80% of cases had at least one listed condition but  INSERT xx% of controls also did.

As shown  in the age band specific univariate logistic regression models in tables \ref{tab:listed.conditions.lt60} to \ref{tab:listed.conditions.ge75}.  Here we collapsed the two younger age bands to <60 years.  All of the listed conditions were significantly associated with severe disease  in each age band. Diabetes was associated with a 6.6. fold rate of severe COVID-19 under age 60 years.  Of note is that "other heart disease" was more  strongly associated with severe COVID-19 than ischaemic heart disease.

Tables \ref{tab:listed.conditions.lt60} to \ref{tab:listed.conditions.ge75}  also show  multivariate rate ratios by age group. The key point of this  analysis is to demonstrate that at all ages there was a very strong  and highly significant  association of having had any prescription in the last year  and of having had any admission in the past five years even adjusted for all the listed conditions and care home residence.

## Systematic analysis of diagnoses associated with severe disease
The approximately two-fold  risk of  with severe  COVID-19  associated with prior hospital admission independently  of other conditions and factors was further explored. The association of hospital diagnoses of each ICD-10 chapter with severe COVID-19, restricted  to those without any of the listed conditions.  These results are shown in Table \ref{tab:conditions}.  In univariate analyses, almost all ICD chapters with at least one diagnosis, with the exception of Chapters 8 (ear) and Chapter 15 (pregnancy) were associated with increased risk of severe disease.  In a multivariate analysis of all chapters, the most strongly associated chapter-level diagnosis was Chapter 2 (neoplasms).  This was further disaggregated with subchapters in as shown in Supplement  Table \ref{tab:icdchapter2}. Unexpectedly, the strongest association was with the subchapter "Malignant Neoplasms of Digestive Organs".  Several other subchapters corresponding to cancer sites were also associated with severe disease.

## Associations of prescribed drugs with severe disease 
Tables \ref{tab:listed.conditions.lt60} to \ref{tab:listed.conditions.ge75} show that the strongest risk factor for severe disease, apart from residence in a care home, is the encashment of any prescription in the last year.  The rate ratio associated with this variable varies from INSERT 9.2 in those aged under 60 years to INSERT 42.2 in those aged 75 years and over.  In a multivariate analysis adjusting for care home residence, any hospital admission and listed conditions, these rate ratios were reduced to  INSERT3.8 and INSERT11.9 respectively.  It is notable that about one third of those over 75 had not encashed a prescription in the previous year.

To investigate this further, we partitioned the "Any prescription" variable into indicator variables for each chapter of the British National Formulary, in which drugs are grouped by broad indication, and restricted the analysis to those without one of the listed conditions.  Table \ref{tab:drugs} shows these associations.  In univariate analyses, prescriptions in almost all BNF chapters were associated with severe disease.  In a multivariate analysis of all chapters, these associations were much weaker.  The BNF chapter with the strongest association with severe disease was Chapter 1 (gastrointestinal).  Other chapters associated with severe disease were Chapter 2 (cardiovascular) and Chapter 4 (central nervous system).

## Construction of a multivariate risk prediction model 
Predictive models were constructed from two sets of variables: a baseline set consisting only of care home residence plus the listed conditions, and an extended set that augmented the baseline set with an indicator variable for ICD-10 chapter level admissions, and indicator variables for prescriptions in each of the BNF chapters.

For each of variable set, a stepwise regression procedure was carried out using alternating forward-backward selection. The variables retained with each variable set are shown in Table \ref{tab:stepwise} . The predictive performance of the model chosen by stepwise regression was estimated by `r nfold`-fold cross-validation.  Observed and predicted case status were compared over all test folds.  Table \ref{tab:xvalid} shows that using the extended set increased the C-statistic from `r table.densities[1, 3]` to `r table.densities[2, 3]` and the expected information for discrimination from `r table.densities[1, 5]` bits to `r table.densities[2, 5] ` bits.

This value for the information conditional on age and sex can be added to the information for discrimination obtained with a logistic regression model using age and sex to obtain the total information for discrimination that would be obtained in the population  to give a total of WHAT??.  PAUL HOW MUCH IS THAT THIS IS NOT CLEAR

Figure \ref{fig:wevid} shows the class-conditional distributions of the weight of evidence favouring case over <!--thi figure should go into a supplement--> control status from the model based on the extended variable set.  This shows, as expected for a multifactorial classifier, that the distributions are approximately Gaussian: there is no clear divide between high-risk and low-risk individuals of the same age and sex.

# Discussion  <!--helen still to edit -->
## Multifactorial risk
In this analysis of sociodemographic factors and past medical history we found a large number of factors to contribute to risk of severe COVID-19 disease.  Apart from the very steep rise in risk with age, severe disease is more likely in men, and those who are more socioeconomically disadvantaged. Being in a residential care home was associated with a 15-fold increased risk of severe COVID-19 in this age matched analysis.

In contrast reports from the U.S and elsewhere in the UK [@the_opensafely_collaborative_opensafely_2020] we found no evidence of an increased risk in non-White minorities. This suggests that ethnic differences found elsewhere are not genetically driven but reflect other social factors. In a recent report based on the UK Biobank cohort, almost half of the fourfold risk ratio for hospitalised COVID associated with Black ethnicity was accounted for by the health care worker status [@niedzwiedz_ethnic_2020;@ho_modifiable_2020 ].

Media reports of apparently otherwise healthy youths succumbing to severe COVID-19 can lead to an impression that severe COVID-19 is a random event. However we found that almost half of cases under 40 years had at least one condition listed by public health agencies as increasing risk.  In those that did not have a listed condition  having any hospital admission  in the past five years and having any prescription in the past year  was much more common  among cases that in the background population of the same age.
